### 2020-10-22 latex/vlp.tex
First edit of the new proof of VLP energy injection

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
   .
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
   .
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
   .
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
   .
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we get the Hamiltonian (\ref{eqn:simply-actuated-hamiltonian}) whose
dynamics (\ref{eqn:simply-actuated-dynamics}) resolve to  
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q)
      , \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau
        . \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \mathbb{R}\) affecting the dynamics of
\(p_l\), acting colinearly with the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This makes the VLP into an underactuated mechanical system with degree of
underactuation 1. 
It is also a useful assumption because it means \((\mathbf{q},\mathbf{p})\) 
are simply actuated coordinates, which allows us to apply the theory of VNHCs we
developed in Chapter \ref{ch:vnhcs}.

Let us define the VNHC \(l = L(q,p)\) -- by which we mean we are
actually defining the VNHC \(h(\mathbf{q},\mathbf{p}) = l - L(q,p) = 0\) of
order 1.
The VLP satisfies
\(\nabla_q \Minv(\mathbf{q}) = \Zmat{2 \times 2}\);
therefore, \(h(\mathbf{q},\mathbf{p})\) is a regular VNHC whenever 
\(L(q,p)\) is \(C^2\) because
\[
   dh_{\mathbf{q}} \Minv(\mathbf{q})B = 
   \begin{bmatrix}
      -\pdiff{L}{q} & 1
   \end{bmatrix}
   \begin{bmatrix}
      \frac{1}{ml^2}  & 0 \\
      0 & \frac{1}{m}
   \end{bmatrix} 
   \begin{bmatrix}
      0 \\ 
      1
   \end{bmatrix}
   = \frac{1}{m}
\]
is always full rank.
The constraint manifold is
\(\Gamma = \mathbb{S}^1 \times \R\), which is parameterized by the unactuated
phase \((q,p)\).
The constrained dynamics are Hamiltonian, and Theorem \ref{thm:zero-dynamics}
tells us they are described entirely by \((\dot{q},\dot{p})\) with \(l\)
replaced by \(L(q,p)\),
\begin{align}\label{eqn:vlp-hamiltonian}
   \mathcal{H}(q,p) &= \frac{1}{2} \frac{p^2}{mL^2} - \frac{1}{2}\dot{L}^2 -
      mgL\cos(q)
   , \\
  &\begin{cases}
      \dot{q} = \frac{p}{m L^2} \\
      \dot{p} = -mgL\sin(q)
      . \\ 
   \end{cases} \nonumber
\end{align}
Note that we suppress the function notation of \(L(q,p)\) and
\(\dot{L}(q,p)\) for clarity.

@@

Because \(L_\frac{\pi}{2}(\theta)\) is smooth and it approximates
\(L^\star(\theta)\), we set our VNHC to be
\[
   h(\mathbf{q},\mathbf{p}) = l - L_\frac{\pi}{2}\left(\theta(q,p)\right)
   .
\]
What remains is to prove this injects energy into the VLP. 
As part of the proof, we will require the following lemma.

\begin{lemma}\label{lemma:sign-of-cube}
   For any \(x,y \in \R\),
   \[
      \sign{x^3 - y^3} = \sign{x-y}
   \]
\end{lemma}
\begin{proof}
   Observe that \(x^3 - y^3 =  (x-y)(x^2 + xy + y^2)\).
   The inequality \(x^2 + xy + y^2 \geq 0\) holds because
   \[
      x^2 + xy + y^2 = \left(x  + \frac{y}{2}\right)^2 + \frac{3y^2}{4} \geq 0
      ,
   \]
   which proves the lemma.
\end{proof}

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   Suppose the initial condition of the VLP's unactuated variables is
   \((q_0,p_0) \not \in \left\{(0,0),(\pm\pi,0)\right\}\).
   A VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) injects energy into the VLP
   if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at the
   coordinate axes.
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
   Let \(\Gamma\) be the constraint manifold for \(h\).
   In negative-time, the dynamics on \(\Gamma\) are
   \begin{equation}\label{eqn:negative-time-vlp}
      f(q,p) = 
      -\begin{bmatrix}
         \dot{q} \\ \dot{p}
      \end{bmatrix} = \begin{bmatrix}
         -\frac{p}{mL(\theta)^2} \\
         mgL(\theta)\sin(q)
      \end{bmatrix}
      .
   \end{equation}
   Let \(D = \Gamma \backslash \{(\pm \pi, 0)\}\)
   be the domain of interest on which we study this negative-time system.
   Choose, as a candidate Lyapunov-like function, the energy for the
   average-length pendulum with zero-potential at the bottom of the swing:
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
   \]
   This is positive definite at \((0,0)\) and has compact sublevel
   sets on \(D\). 
   The derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \[
      \dot{E}_\text{avg} = \frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
              {l_\text{avg}^2L(\theta)^2}
      .
   \]
   The sign of \(\sin(q)p\) is dependent only on the quadrant in the
   \((q,p)\)-plane, so one can show it is equal to the sign of
   \(\sin(2\theta)\). 
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{l_\text{avg}^3 - L(\theta)^3} = \sign{l_\text{avg} - L(\theta)}
       .
   \]
   The derivative of \(E_\text{avg}\) is nonnegative since
   \begin{align*}
    \sign{\dot{E}_\text{avg}} &= \sign{\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)} \\
                &= \sign{\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)} \\
                &= -\sign{(l_\text{avg} - L(\theta))\sin(2\theta)} \\
                &\leq 0 \text{ (by assumption)}
      .
   \end{align*}
   Thus, \(E_\text{avg}\) is a Lyapunov function and the negative-time system is
   stable \cite{lyapunov}. 
   The last step of the proof is to show that (\ref{eqn:negative-time-vlp})
   asymptotically converges to \((0,0)\). 
   For this, we will use a Krasovskii-LaSalle argument
   \cite{krasovskii_lasalle}.
   First, we define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\).
   Since \(f(q,p) = (0,0)\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the VLP is gaining energy on average.
   Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
   means \(E_\text{avg}\) is a Lyapunov function for the positive time system;
   by the same arguments presented here, the VLP loses energy while converging
   asymptotically to \((0,0)\).
 \end{proof}

>>>>

Using this configuration, we will compute the Hamiltonian dynamics of the system.
The cartesian position of the mass at the tip of the pendulum
is given by \(x = (l\sin(q),-l\cos(q))\), while its velocity is
\(\dot{x} = (\dot{l}\sin(q) + l\cos(q)\dot{q}, -\dot{l}\cos(q) + l\sin(q)\dot{q})\).
Computing the kinetic energy \(T\) yields
\[
   T(\mathbf{q},\dot{\mathbf{q}}) = 
   \frac{1}{2}m\norm{\dot{x}}^2 = \frac{1}{2}m\left(\dot{l}^2 + l^2\dot{q}^2\right)
   .
\]
The potential energy \(P\) with respect to the pivot (under a gravitational
acceleration \(g\)) is
\[
   P(\mathbf{q}) = -mgl\cos(q)
   .
\]
Collecting the kinetic energy into a quadratic form, we get the Lagrangian
\[
   \mathcal{L}(\mathbf{q},\dot{\mathbf{q}}) 
   = \frac{1}{2} \dot{\mathbf{q}}\tpose D(\mathbf{q})\dot{\mathbf{q}} - P(\mathbf{q})
   = \frac{1}{2}
   \begin{bmatrix} \dot{q} & \dot{l} \end{bmatrix}
   \begin{bmatrix}
      ml^2 & 0 \\
      0 & m \\
   \end{bmatrix}
   \begin{bmatrix} 
      \dot{q} \\ \dot{l}
   \end{bmatrix}
   + mgl\cos(q)
   .
\]
Computing the conjugate of momenta to \(\mathbf{q}\), we get 
\[
   \mathbf{p} := \begin{bmatrix} p \\ p_l \end{bmatrix} 
   = \begin{bmatrix} ml^2\dot{q} \\ m\dot{l} \end{bmatrix} 
   .
\]
Performing the Legendre transform on \(\mathcal{L}\) and setting
\(M(\mathbf{q}) := D(\mathbf{q})\), \(V(\mathbf{q}) := P(\mathbf{q})\),
we get the Hamiltonian (\ref{eqn:simply-actuated-hamiltonian}) whose
dynamics (\ref{eqn:simply-actuated-dynamics}) resolve to  
\begin{align}\label{eqn:vlp-hamiltonian-with-pl}
   \mathcal{H} &= \frac{1}{2} \begin{bmatrix} p & p_l \end{bmatrix}
      \begin{bmatrix}
         \frac{1}{ml^2}  & 0 \\
         0 & \frac{1}{m}
      \end{bmatrix} \begin{bmatrix} p \\ p_l \end{bmatrix} - mgl\cos(q)
      , \\
     &\begin{cases}
        \dot{q} = \frac{p}{ml^2} \\
        \dot{l} = \frac{p_l}{m} \\
        \dot{p} = -mgl\sin(q) \\
        \dot{p}_l = \frac{p^2}{ml^3} + mg\cos(q) + \tau
        . \\
   \end{cases} \nonumber
\end{align}
The control input is a force \(\tau \in \mathbb{R}\) affecting the dynamics of
\(p_l\), acting colinearly with the rod.
We assume the force does not affect the dynamics of \(p\) in any way -
that is, the control input cannot enact any lateral force on the pendulum.
This makes the VLP into an underactuated mechanical system with degree of
underactuation one. 
It is also a useful assumption because it means \((\mathbf{q},\mathbf{p})\) 
are simply actuated coordinates, which allows us to apply the theory of VNHCs we
developed in Chapter \ref{ch:vnhcs}.

Let us define the VNHC \(l = L(q,p)\), by which we mean we are
actually defining the VNHC \(h(\mathbf{q},\mathbf{p}) = l - L(q,p) = 0\) of
order 1.
The VLP satisfies
\(\nabla_q \Minv(\mathbf{q}) = \Zmat{2 \times 2}\).
By Theorem \ref{thm:vnhc-regularity},
\(h(\mathbf{q},\mathbf{p})\) is a regular VNHC whenever 
\(L(q,p)\) is \(C^2\) because
\[
   dh_{\mathbf{q}} \Minv(\mathbf{q})B = 
   \begin{bmatrix}
      -\pdiff{L}{q} & 1
   \end{bmatrix}
   \begin{bmatrix}
      \frac{1}{ml^2}  & 0 \\
      0 & \frac{1}{m}
   \end{bmatrix} 
   \begin{bmatrix}
      0 \\ 
      1
   \end{bmatrix}
   = \frac{1}{m}
\]
is always full rank.
The constraint manifold is
\(\Gamma = \mathbb{S}^1 \times \R\), which is parameterized by the unactuated
phase \((q,p)\).
The constrained dynamics are Hamiltonian, and Theorem \ref{thm:zero-dynamics}
tells us they are described entirely by \((\dot{q},\dot{p})\) with \(l\)
replaced by \(L(q,p)\):
\begin{align}\label{eqn:vlp-hamiltonian}
   \mathcal{H}(q,p) &= \frac{1}{2} \frac{p^2}{mL^2} - \frac{1}{2}\dot{L}^2 -
      mgL\cos(q)
   , \\
  &\begin{cases}
      \dot{q} = \frac{p}{m L^2} \\
      \dot{p} = -mgL\sin(q)
      . \\ 
   \end{cases} \nonumber
\end{align}
Note that we suppress the function notation of \(L(q,p)\) and
\(\dot{L}(q,p)\) for clarity.

@@

Because \(L_\frac{\pi}{2}(\theta)\) is smooth and it approximates
\(L^\star(\theta)\), we set our VNHC to be
\[
   h(\mathbf{q},\mathbf{p}) = l - L_\frac{\pi}{2}\left(\theta(q,p)\right)
   .
\]

Before continuing, we must rigorously define what it means for a VNHC to inject
energy into the VLP.
We want the the VLP to behave in
such a way that momentum increases over time;
this intuition is captured mathematically by the following definitions.

\begin{defn}\label{defn:energy-injection}
   Let \(\dot{x} = f(x)\) be a planar ODE defined on \(D \subset \R^2\), with
   \(f(0) = (0,0)\).
   This system \textit{gains energy on \(D\)} if 
   \begin{enumerate}
      \item The origin is a repeller 
         (or equivalently, the origin is asymptotically stable in negative time).
      \item For all compact sets \(K \subset D\) and for almost every initial
         condition \(x(0) \in K\), there exists \(T > 0\) such
         that \(x(t) \not \in K \, (\forall t > T)\).
   \end{enumerate}
\end{defn}

\begin{defn}
   A regular VNHC \(h(\mathbf{q},\mathbf{p}) = 0\) 
   \textit{injects energy into the VLP} if the constrained dynamics gain energy
   everywhere on the constraint manifold, except possibly on a set of measure
   zero.
\end{defn}

We can now prove our VNHC injects energy into the VLP. 
As part of the proof, we will require the following lemma.

\begin{lemma}\label{lemma:sign-of-cube}
   For any \(x,y \in \R\),
   \[
      \sign{x^3 - y^3} = \sign{x-y}
      .
   \]
\end{lemma}
\begin{proof}
   Observe that \(x^3 - y^3 =  (x-y)(x^2 + xy + y^2)\).
   The inequality \(x^2 + xy + y^2 \geq 0\) holds because
   \[
      x^2 + xy + y^2 = \left(x  + \frac{y}{2}\right)^2 + \frac{3y^2}{4} \geq 0
      ,
   \]
   which proves the lemma.
\end{proof}

@@

\begin{thm}\label{thm:vlp-energy-stabilization}
   Define \(\theta := \arctan_2(p,q)\).
   Let \(L : \mathbb{S}^1 \rightarrow [\underline{l},\overline{l}]\) be a
   \(C^2\) function of \(\theta\).
   A regular VNHC of the form 
   \(h(\mathbf{q},\mathbf{p}) = l - L(\theta)\) injects energy into the VLP
   if there exists 
   \(l_\text{avg} \in [\underline{l},\overline{l}]\) such that 
   \begin{equation}\label{eqn:vlp-energy-gain-condition}
      \left(l_\text{avg} - L(\theta)\right)sin(2\theta) \geq 0 
      \text{ }\forall \theta \in \mathbb{S}^1
      ,
   \end{equation}
   with the property that the inequality is strict except at the
   coordinate axes.
   If the inequality is flipped, the VNHC dissipates energy. 
\end{thm}
\begin{proof}
   Let \(\Gamma = \mathbb{S}^1 \times \R \) be the constraint manifold for \(h\).
   In negative-time, the dynamics on \(\Gamma\) are
   \begin{equation}\label{eqn:negative-time-vlp}
      f(q,p) = 
      -\begin{bmatrix}
         \dot{q} \\ \dot{p}
      \end{bmatrix} = \begin{bmatrix}
         -\frac{p}{mL(\theta)^2} \\
         mgL(\theta)\sin(q)
      \end{bmatrix}
      .
   \end{equation}
   We begin by analyzing the upright equilibria \((\pm \pi, 0)\). 
   The linearization of (\ref{eqn:negative-time-vlp}) at these
   equilibria is given by
   \[
      df_{(\pm \pi,0)} = \begin{bmatrix}
         0 & -\frac{1}{mL(\pi)^2} \\
         -m g L(\pi) & 0 \\
      \end{bmatrix}
      ,
   \]
   which has eigenvalues at \(\lambda = \pm \sqrt{\frac{g}{L(\pi)}}\).
   The positive-time system has the same eigenvalues at these upright
   equilibria.
   We label the set of initial conditions converging to \((\pm \pi, 0)\) in
   positive time by
   \[
      \Pi_0^+ := \left\{ (q_0,p_0) \in \Gamma \mid \lim_{t \to \infty} (q(t),p(t)) = (\pm \pi, 0) \right\}
      ,
   \]
   and the set of initial conditions converging to \((\pm \pi,0)\)
   in negative time by \(\Pi_0^-\).
   Since only one eigenvalue is negative, the stable manifold theorem tells us
   that both \(\Pi_0^+\) and \(\Pi_0^-\) are one-dimensional \textbf{SOURCE?},
   which means they have measure zero within \(\Gamma\).

   To prove the origin of the positive-time system is a repeller, we will show
   the origin of the negative-time system is almost globally
   asymptotically stable.

   Choose, as a candidate Lyapunov function, the energy for the
   average-length pendulum with zero-potential at the bottom of the swing:
   \[
       E_\text{avg}(q,p) := \frac{1}{2}\frac{p^2}{m l_\text{avg}^2} 
           + m g l_\text{avg} (1-\cos(q))
      .
   \]
   This is positive definite at \((0,0)\) and has compact sublevel
   sets on \(\Gamma\). 
   The derivative of \(E_\text{avg}\) along (\ref{eqn:negative-time-vlp}) is
   \[
      \dot{E}_\text{avg} = \frac{g\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)}
              {l_\text{avg}^2L(\theta)^2}
      .
   \]
   The sign of \(\sin(q)p\) depends only on the quadrant in the
   \((q,p)\)-plane. In fact, it is easy to show that
   \[
      \sign{\sin(q)p} = \sign{\sin(2\theta)}
      .
   \]
   Furthermore, by Lemma \ref{lemma:sign-of-cube} we have
   \[ 
       \sign{L(\theta)^3 - l_\text{avg}^3} = \sign{L(\theta) - l_\text{avg}}
       .
   \]
   The derivative of \(E_\text{avg}\) is never positive since
   \begin{align}\label{eqn:vlp-proof-neg-inv}
    \sign{\dot{E}_\text{avg}} &= \sign{\sin(q)p \left(L(\theta)^3 - l_\text{avg}^3\right)} \\
                &= \sign{\sin(2\theta) \left(L(\theta) - l_\text{avg}\right)} \\
                &= -\sign{(l_\text{avg} - L(\theta))\sin(2\theta)} \\
                &\leq 0 \text{ (by assumption)}
      .
   \end{align}
   Thus, \(E_\text{avg}\) is a Lyapunov function, which means the negative-time system is
   stable \cite{lyapunov}. 
   Define \(D := \Gamma \backslash \Pi_0^-\) to be the domain of interest.
   Using a Krasovskii-LaSalle argument \cite{krasovskii_lasalle}, 
   we will prove that (\ref{eqn:negative-time-vlp}) asymptotically converges to
   \((0,0)\) if it is initialized on \(D\).
   First, we define the zero-set
   \[
      Z = \left\{(q,p) \in \Gamma \mid 
      \dot{E}_\text{avg}(q,p) = 0 \right\}
      .
   \]
   It is easy to show that \(Z\) is the union of the coordinate axes
   \[
      Z= \mathbb{S}^1 \times \left\{0\right\} \cup 
         \left\{0\right\} \times \R
      .
   \]
   Krasovskii-LaSalle tells us that solutions to (\ref{eqn:negative-time-vlp})
   converge to the largest invariant set in \(Z \cap D\).
   Since \(f(q,p) = (0,0)\) only when \((q,p) = (0,0)\),
   the origin is asymptotically stable on \(D\) for the negative-time system.
   Hence, the positive-time system is diverging away from the origin 
   in the \((q,p)\) plane, which means the origin is a repeller.

   We now prove the system escapes compact sets in finite time.
   Let \(x(t) = (q(t),p(t))\) be the orbit for the positive-time system.
   Let \(K \subset \Gamma\) be compact and assume \(x(0) \in K\backslash \Pi_0^+\).
   For a proof by contradiction, suppose that for all \(T > 0\) there exists \(t > T\)
   where \(x(t) \in K\).
   Let 
   \[
      k := \max\limits_{(q,p) \in K} E_\text{avg}(q,p)
      ,
   \] 
   be the largest energy value attained in \(K\), and let
   \[
      E_k := \left\{ (q,p) \in \Gamma \mid E_\text{avg}(q,p) \leq k \right\}
      ,
   \]
   be the energy sublevel set with that largest energy. 
   Note that \(E_k\) is compact, contains \(K\), and is negatively invariant (it
   is positively invariant for the negative-time system).
   Because \(K \subset E_k\), it follows that \(x(t) \in E_k\). 
   Then, since \(E_k\) is negatively invariant, the half orbit
   \(x([0,t])\) is contained fully within \(E_k\).
   In particular, \(x(T) \in E_k\); 
   but \(T\) was chosen arbitrarily, which implies \(x(t) \in E_k\) for all
   \(t \in [0,\infty[\).
   Since \(E_k\) is compact, \(x(t)\) has a positive limit set in
   \(E_k\) which (by Poincar\'{e}-Bendixson) is either an equilibrium or a
   closed orbit \textbf{SOURCE?}.
   The system cannot have closed orbits, since this would violate asymptotic
   stability of \((0,0)\) in negative time. 
   Hence, the positive limit set must be one of the equilibria. 
   The origin is a repeller, which means \(x(t)\) must be converging to 
   \((\pm \pi,0)\).
   This is a contradiction, since we assumed \(x(0) \not \in \Pi_0^+\).

   We have shown that orbits of the constrained dynamics will leave compact sets in
   finite time, including all sublevel sets of \(E_\text{avg}\). 
   The orbit \((q(t),p(t))\) is spiraling away from the origin and (for almost
   all initial conditions) will not hit the upper equilibrium.
   Thus, the VLP is gaining energy on average because its momentum is increasing.
   Flipping the inequality of (\ref{eqn:vlp-energy-gain-condition}) 
   means \(E_\text{avg}\) is a Lyapunov function for the positive time system;
   by the Krasovskii-LaSalle arguments presented earlier, the VLP loses energy
   while converging asymptotically to \((0,0)\).
 \end{proof}
